---
author: Whitney Friedman
date: 2/18/2020
title: "SNAPP Postdoc Training"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started with R Markdown

Some useful R markdown commands: 
**Bold text** 

Lists

* apple
* banana
* kiwi

Need to include a blank line in code before a new paragraph. 

Help > Markdown Quick Reference

* Basic commands
* Can use simple LaTeX code (for equations)
* $\sum{x^2}$

Help > Cheatsheets

* Downloadable cheatsheets here
* See: Rmarkdown reference guide

# Open a new R chunk
* CMD-Option-i to open new code chunk (default: R)
* RMD also supports python and bash command
* Note, when you knit, it opens code in a new environment. Variables run there are not saved locally.
* But if you hit the "play" button, it runs the code locally (and saves variables here)
* Gear icon > chunk output inline  / chunk output in console.
* Gear > clear all output
* Naming chunks is a style thing. Each one needs a distinct name within a document. 

```{r new-chunk}
# Type normal r-code here; print x
x <- c(1,13,-5)
x
```


```{r}
y <- x + 4
print(y)
```

# Chunk options
* If you set echo = False, the chunk is RUN but its hidden from the knit document. Outputs are shown.
* If you set eval = False, the chunk will not be run
* message / warning = False, don't show warnings... 
* See more at help > cheatsheets > R markdown reference guide

```{r, echo=FALSE}
y**3
```

# This chunk won't be run. 
```{r eval = FALSE}
print("this chunk is not evaluated")
```

```{r}
# test above
```



