---
title: "01-DataCleaning"
author: "Whitney"
date: "2/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data cleaning
Following Ch10 in https://science-for-nature-and-people.github.io/2020-data-collab-workshop/2020-02-snapp/data-cleaning-and-manipulation.html

## Load libraries
Good to always do this first, at the top of your code.
Lets others know what the dependencies are.
Can include an if-statement to install if needed (if you're sharing across machines/people)

e.g. 

packages <- c("DT", "devtools", "tidyverse", "ggmap", "ggplot2", "leaflet", "readxl", "tidyr", "scales", "sf", "raster", "rmarkdown", "roxygen2", "broom", "captioner")

for (package in packages) { if (!(package %in% installed.packages())) { install.packages(package) } }

If you don't want to see library warnings everytime (after you read it!), set warning=F, message=F

```{r load-libraries, warning= F, message = F}
library(dplyr)
library(tidyr)
```

## Read in data
We are using data from
[Mike Byerly. Alaska commercial salmon catches by management region (1886- 1997). Gulf of Alaska Data Portal. df35b.304.2](https://knb.ecoinformatics.org/view/df35b.304.2)

(See above for hyperlink formatting in RMD)

Keep 'original' version of dataframe around without having to read it in again. 

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1",method="libcurl"),stringsAsFactors = F)

head(catch_original)

```

## Clean and reshape
Remove marginal sum and notes columns using `select()`

Note: The pipe operator `%>%` (cmd-shift-m). Read as "then". More computationally efficient than rows of re-assignment. 

Warning: NAs introduced by coercion. Check: what did I assume? 

```{r}
catch_clean <- catch_original %>%
  select(Region,Year, Chinook, Sockeye, Coho, Pink, Chum) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% # else = keep original value of Chinook.
  mutate(Chinook = as.numeric(Chinook))

# Review data
head(catch_clean)
summary(catch_clean) # Chinook is character
```

### QA
Find where as.numeric couldn't convert chinook to a number.

```{r}
i <- which(is.na(catch_clean$Chinook))
i
```
```{r}
catch_original[i,]
```

### Reshape wide to tall
Using (new) function `pivot_longer`

* `melt`/`cast` - from plyr
* `gather`/`spread` - from dplyr (updated from plyr::melt/cast)
* `pivot_wider`/`pivot_longer` - updated from dplyr::gather/spread

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region,Year), 
               names_to = "species",
               values_to = "catch") %>% 
  #mutate(catch = catch * 1000) %>% # overwrite then rename to not create a new column
  #rename(catch_thousands = "catch" ) %>% # rename: value new = value old
  mutate(catch_thousands = catch * 1000)# new column
  

head(catch_long)

```
### Example with `pivot_wider`

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = Year,
              values_from = catch)

head(catch_wide)
```
### group_by and summarize

group_by: on the backend, adds grouping information to dataframe. 

```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarise(catch_mean = mean(catch),
            num_obs = n()) # count number of observations in each group

head(mean_region)
```

```{r}
mean_species <- catch_long %>% 
  group_by(species) %>% 
  summarise(mean_catch = mean(catch),
            num_obs = n())

head(mean_species)
```
```{r}
# in what year was the total salmon catch highest? 
annual_catch <- catch_long %>% 
  group_by(Year) %>% 
  summarise(total_catch = sum(catch)) %>% 
  arrange(-total_catch)

annual_catch

```

```{r}
# what year had the highest sockeye catch?
# using "filter" 
annual_sockeye <- catch_long %>% 
  filter(species == "Sockeye") %>% 
  group_by(Year) %>%
  summarise(total_catch = sum(catch)) %>% 
  arrange(-total_catch)

annual_sockeye
```

## Joins 

See [Ch 10.14 "Joins in dplyr"](https://science-for-nature-and-people.github.io/2020-data-collab-workshop/2020-02-snapp/data-cleaning-and-manipulation.html#joins-in-dplyr)

Think about how many rows and columns you expect to return before doing the join. 
e.g. If I get as many rows returned as I have in my left table, was correct. 
if RHS table has duplicate keys, then the result of a left_join will be twice as long. 
(This indicates a problem with the original data).

```{r}
# Want to join the name of the management area to our df. 
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1",method="libcurl"),stringsAsFactors = F) %>% 
  select(code,mgmtArea) # can select at the time of read. 
  
region_defs
```
```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code")) # c(LHS key = RHS) key

# Equivalent, but less explicit: 
#catch_joined <- catch_long %>% 
#  left_join(region_defs)

catch_joined
```




## Join to regions table