---
title: "02-data-vis"
author: "Whitney"
date: "2/20/2020"
output: html_document
---

# Data Visualization 
Following along from 
[Ch 12, SNAPP Book](https://science-for-nature-and-people.github.io/2020-data-collab-workshop/2020-02-snapp/publication-graphics.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages
```{r message=FALSE, warning=FALSE}

library(leaflet)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales) # install.packages("scales")
library(patchwork) # install.packages("patchwork")

```

## Load data
```{r}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

# If the file exists locally, read it, otherwise get it from KNB
esc <- tryCatch(
    read.csv("data/escapement.csv", stringsAsFactors = FALSE),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"), stringsAsFactors = FALSE)
        return(esc)
    }
)

head(esc)

```

## Data processing

```{r}
annual_esc <- esc %>% 
  # Split column (of character data), default drops the original column.
  separate(sampleDate, c("Year","Month", "Day"), sep = "-") %>% 
  mutate(Year = as.numeric(Year),
         Month = as.numeric(Month),
         Day = as.numeric(Day)) %>% 
  group_by(Species, SASAP.Region, Year) %>% 
  summarise(escapement = sum(DailyCount)) %>% 
  filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))
  
head(annual_esc)
```
```{r}

ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col(fill = "blue")

```

## Per region

```{r}
ggplot(annual_esc, aes(x=Species, y=escapement, fill =  SASAP.Region)) + 
  geom_col()
```
### Looking into the Kodiak region 
By adding filter to ggplot 

Hint! the `<TAB>` key will show options in things like `theme()`

Jeanette's oder of operations for ggplot: 
* Data
* GEOMS
* Scaling
* Titles / labels
* Themes

```{r}
annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% 
  ggplot(aes(x = Year, y  = escapement, color=Species)) + 
  # Any GEOM's
  geom_line() + geom_point() + 
  # Any scaling
  # Add titles / labels
  ggtitle("Salmon Escapement in the Kodiak Region") + 
  ylab("Escapement") + 
  # Any theme calls. Adjust almost anything about the appearance with 'theme'
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") # Center the title. 

```

### Save a theme as an object:

```{r}
my_theme <- theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + 
  theme(legend.title = element_blank()) # Use 'element_blank' to delete things
```

Try it out. 

```{r}
annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% 
  ggplot(aes(x = Year, y  = escapement, color=Species)) + 
  # Any GEOM's
  geom_line() + geom_point() + 
  # Any scaling
  # Add titles / labels
  ggtitle("Salmon Escapement in the Kodiak Region") + 
  ylab("Escapement") + 
  my_theme 

```

## Playing with Facet

```{r}

annual_esc %>% 
  ggplot(aes(x = Year, y  = escapement, color=Species)) + 
  geom_line() + geom_point() + 
  scale_y_continuous(labels=comma)+ # change the way numbers are shown
  facet_wrap(~SASAP.Region, scales="free_y",ncol=2) + # free_y allows y to scale to data
  ggtitle("Salmon Escapement in the Kodiak Region") + 
  ylab("Escapement") + 
  my_theme 
```
## Patchwork (package)
A nice package to create multi panel figures!

```{r}
library(package)

# Save plot into a variable (its just a list)
p1 <- ggplot(annual_esc, aes(x=Species, y=escapement)) + 
  geom_col() 
  
p2 <- ggplot(annual_esc, aes(x=Species, y=escapement)) + 
  geom_col(fill = "blue")

p1 + p2 # Use patchwork to make one figure
```


